Parametricity Consequences of GADTs Representations

### 2021-02-14 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_a$ of the first-order functor
whose action on objects is given by $F_a\,X = 1 + a \times X$ and,
whose action on morphisms (which models its corresponding
$\mathsf{map}$ function) is given by $F_a\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is then modelled by the
higher-order functor $H$ whose action on objects is given by $H\,F\,X
= 1 + X \times H\,F\,X$ and whose action on morphisms is givne by
$H\,F\,f = \id_1 + f \times H\,F\,f$. (A {\em higher-order functor} in
this context is a functor not directly on the category interpreting
types, but rather on the category of functors interpreting type
constructors, i.e., on the category of functors on the category of
functors interpreting types.) The ADT $\mathsf{List\,a}$ is then
modelled as the functor $(\mu H)a$. Since $H$ is a higher-order
functor, not only does it have a corresponding functorial action,
i.e., support a function $\mathsf{map_H}$, but $\mu H$ is also a
functor with its own functorial action. That is, the type constructor
$\mathsf{List}$ itself supports a function $\mathsf{fmap_{List}}$.

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., the syntactic reflection of semantic equality. Such a
type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
%, in a higher-kinded calculus such as $F\omega$,
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] 
Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$, e.g., can be written as 
{\color{red} START HERE}
\[ {\color{blue} WHAT GOES HERE??} \]

Note that neither Eq nor Seq has a syntactic map (because one for Eq
leads to a contradition as in Daniel's exampe below, and because one
for Seq can't exist because of Pair), i.e., is not an instance of the
Functor class, and is not required to in this representation.

data Equal a b where
  Refl :: Equal a a

eqElim :: Equal a b -> a -> b
eqElim Refl x = x

emptyFromMapEq :: (forall a b c. (a -> b) -> Equal a c -> Equal b c) -> Void
emptyFromMapEq fmap = eqElim (fmap absurd Refl) ()

@@

Moreover, because its type argument varies in the instances of
$\mathsf{Seq}$ appearing in the types of its data constructors,
$\mathsf{Seq}$ cannot be characterized by any first-order fixpoint
equation. It can, be viewed as a solution to the higher-order fixpoint
equation
\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$~\cite{mac71}.

Other GADTs
are similarly characterizable by higher-order functors.  The
difference between ADTs and GADTs can thus be seen as a difference in
{\em orders} of their underlying functors. {\color{blue} The role of
  higher-order functors here is essential.}

{\color{blue} GADTs make essential use of higher-order functors.}

@@

Having interpretations as fixpoints of higher-order functors, nested
types also have corresponding $\mathsf{fmap}$ functions.

@@

>>>>

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_a$ of the first-order functor
whose action on objects is given by $F_a\,X = 1 + a \times X$ and,
whose action on morphisms (which models its corresponding
$\mathsf{map}$ function) is given by $F_a\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is then modelled by the
higher-order functor $H$ whose action on objects is given by $H\,F\,X
= 1 + X \times F\,X$ and whose action on morphisms is given by
$H\,F\,f = \id_1 + f \times F\,f$. (A {\em higher-order functor} in
this context is a functor not directly on the category interpreting
types, but rather on the category of functors interpreting type
constructors, i.e., on the category of functors on the category of
functors interpreting types.) The ADT $\mathsf{List\,a}$ is then
modelled as the functor $(\mu H)a$. Since $H$ is a higher-order
functor, not only does it have a corresponding functorial action,
i.e., support a function $\mathsf{map_H}$, but $\mu H$ is also a
functor with its own functorial action. That is, the type constructor
$\mathsf{List}$ itself supports a function $\mathsf{fmap_{List}}$.

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
$\Eq$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
%, in a higher-kinded calculus such as $F\omega$,
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\]
Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$ can be written as
%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
%  \times Seq\,c} \] 
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
  Seq \,b \to Seq\,c \to Seq\,a}\\ 
\end{array}\]

This rewriting of $\mathsf{Seq}$'s definition shows that it can be
seen as a fixpoint $\mu H$ of the higher-order function $H$ defined on
objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
F\,c$. We note, however, that $Eq\,a$ cannot be made a functor, or,
equivalently, that $\mathsf{Eq\,a}$ cannot be an instance of Haskell's
$\mathsf{Functor}$ class. Indeed, if $\mathsf{Eq\,a}$ supported a
function $\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} wasEq\,a\,b\to b\to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the following element of the type
$\mathsf{Void}$:
\[\begin{array}{l}
\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
  b \to Eq\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
Of course this is impossible.  It follows immediately that
$\mathsf{Seq}$ cannot support a map function either. This is already
evident from the Haskell definition of $\mathsf{Seq}$, though. The
clause of $\mathsf{map}$ for the $\mathsf{Pair}$ constructor should
feed $\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and
a term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$
and $\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$
for some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. The use of {\em higher-order functions} is
essential, since the type arguments to the GADT being defined are not
uniform across all of its instances in the types of its data
constructors; the use of higher-order funct{\em ors} is essential, as
for ADTs, to the very existence of the fixpoints being
computed~\cite{tfca}.

Consider, for example, the GADT $\mathsf{Seq}$ from above.  Because
its type argument varies in the instances of $\mathsf{Seq}$ appearing
in the types of its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$
cannot be characterized by any first-order fixpoint equation. It can,
however, be viewed as a solution to the higher-order fixpoint equation
{\color{blue} syntax or semantics?}
\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$~\cite{mac71}.

Other GADTs are similarly characterizable by higher-order functors.
The difference between ADTs and GADTs can thus be seen as a difference
in {\em orders} of their underlying functors. {\color{blue} GADTs
  make essential use of higher-order functors.}

@@

@@

Functorial comnpletion doesn't give nested types
